Red Wine Analysis by Riyazudeen Abdul Subhan
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gridExtra)
library(GGally)
library(scales)
library(memisc)
library(ggthemes)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
pd <- read.csv('wineQualityReds.csv', sep =',')
```

#Introduction
I have dataset about Redwines and I am planning to explore the data and going to
find out the intresting and hidden facts about red wine :). Wine is one of the 
most common and popular drinks among the people so let's explore our data sample
to find more about it. 

Let's see the strucute of the data 

```{r echo=FALSE, Load_the_Data1}
str(pd)
summary(pd)
```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
#Create function to faciliate plotting of histograms in ggplot
plot.hist <- function(x.var, df, bw){
   (ggplot(df, aes_string(x = x.var)) +
      geom_histogram(binwidth = bw, colour="#FF9999"))
    # geom_histogram(binwidth = bw,stat="identity", colour="#FF9999")
}

# plot.hist(whatever you want) +scale_x_continuous(breaks(seq(0,1,0.2)))

```

```{r echo=FALSE,Univariate_Plots_fixed.acidity}
plot.hist('fixed.acidity', pd, 0.3) + 
  scale_x_continuous(breaks = seq(0, 16, 1)) +
  labs(y = 'Fixed Acidity', x = 'Count',
          title = "Fixed Acidity Distribution") 

with(pd, summary(fixed.acidity))
```


I have noticed that most of wine have a fixed acidity between 5 and 12.

```{r echo=FALSE,Univariate_Plots_volatile.acidity}

plot.hist('volatile.acidity', pd, 0.05) + 
  scale_x_continuous(breaks = seq(0, 1.6, 0.2 )) 

with(pd, summary(volatile.acidity))

```
The wine with the lowest volatile acidity has a score of 0.12 and the highest 
has 1.58

```{r echo=FALSE,Univariate_Plots_citric.acid}
plot.hist('citric.acid', pd, 0.03) +
    scale_x_continuous(breaks = seq(0, 1, 0.1)) 
with(pd, summary(citric.acid))
```
Most values lie between 0.09 and 0.42. The wine with the citric acid level of 1 
is the only outlier in this distribution.

```{r echo=FALSE,Univariate_Plots_residual.sugar}
plot.hist('residual.sugar', pd, 0.1) +
  coord_cartesian(xlim = c(0.2, 4))

with(pd, summary(residual.sugar))

```
Most wines have a residual sugar level value between 1.9 and 2.6. The wine with 
the lowest residual sugar has a value of 0.9 and the highest has a score of 15.5

```{r echo=FALSE,Univariate_Plots_chlorides}
plot.hist('chlorides', pd, .001)

plot.hist('chlorides', pd, .001) +
  coord_cartesian(xlim = c(0.012, 0.2))

with(pd, summary(chlorides))
```
The chloride levels for the wines in the dataset range from 0.012 to 0.611. Most
values lie between 0.07 and 0.09. There are multiple outliers in this 
distribution where the wines have chloride levels higher than 0.13.


```{r echo=FALSE,Univariate_Plots_free.sulfur.dioxide}
plot.hist('free.sulfur.dioxide', pd, 1)

plot.hist('free.sulfur.dioxide', pd, .05) + scale_x_log10()


with(pd, summary(free.sulfur.dioxide))

```

Most wines have values between 7 and 21. The lowest value is 1 and the highest 
is 72. There are multiple outliers in this distribution where the wines have 
free sulfur dioxide levels higher than 42.

```{r echo=FALSE,Univariate_Plots_total.sulfur.dioxide}
plot.hist('total.sulfur.dioxide', pd, 5)

plot.hist('total.sulfur.dioxide', pd, .02) + scale_x_log10()

with(pd, summary(total.sulfur.dioxide))

```

Total sulfur dioxide levels range between 6 and 289. Most values lie between 22 
and 62. 

```{r echo=FALSE,Univariate_Plots_density}
plot.hist('density', pd,0.0001) +
  labs(y = 'Density', x = 'Count',
          title = "Density distribution") 

with(pd, summary(density))
```

The density of the wine's in the dataset range from 0.9901 to 1.004. Most wines 
lie between 0.9956 and 0.9978. 

```{r echo=FALSE,warning=FALSE,Univariate_Plots_pH}
plot.hist('pH', pd, .01) + scale_x_continuous(limits = c(3, 4)) 

with(pd, summary(pH))
```

The pH levels of wine range from 2.74 to 4.01. Most have values between 3.21 and
3.4. There are multiple outliers in this distribution where the wines have pH 
levels that are either higher than 3.6.

```{r echo=FALSE,Univariate_Plots_sulphates}
plot.hist('sulphates', pd, .01)

with(pd, summary(sulphates))

```

Sulphate levels range from 0.33 to 2.00. Most values lie between 0.55 and 0.73. 

```{r echo=FALSE,Univariate_Plots_alcohol}
plot.hist('alcohol', pd, .5) +
  labs(y = 'Alcohol', x = 'Count',
          title = "Alcohol distribution") +
      scale_x_continuous(breaks = seq(0, 14,.5 )) 

with(pd, summary(alcohol))
```

Most wines have alcohol levels between 9.5 and 11.1. The minimum is 8.4 and the 
maximum value is 14.9. 

```{r echo=FALSE,Univariate_Plots_quality}
ggplot(data = pd, aes(x = factor(quality))) + 
  geom_bar()

with(pd, summary(quality))

```

The quality ratings of wine range from 3 to 8. Most wines having a rating of 5 
or 6. 

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains 1599 types of wines each with 11 independent variables that
are based on physicochemical tests and one dependent variable. The dependent 
variable is the measurement of quality which is based on sensory data provided 
by wine experts. Quality variable has maximum value of 8 and min value 3 and 
there is no wine sample in the dataset which has quality of less than 3 and 
greater than 8. 2

### What is/are the main feature(s) of interest in your dataset?

The univariate analysis of the red wine dataset revealed interesting 
distributions in alcohol, fixed acidity, and sulphates. Each of these variables 
has long tails which I will investigate further in my bi-variate analysis. I 
suspect that these variables have an influence on the quality of the wine.

### What other features in the dataset do you think will help support your 
###investigation into your feature(s) of interest?

The other two variables that I will examine in more detail are citric acid and 
total sulfur dioxide. Citric acid is added to finished wines to increase acidity
and give the wine a 'fresh flavor.' If too much is added, it can increase the 
formation of volatile acid which gives wines a vinegar taste. There may be a 
'sweet spot' for this variable which I will explore in more detail. Plotting the
values of total sulfur dioxide revealed a bi modal distribution which may be 
related to the quality of the wine.

### Did you create any new variables from existing variables in the dataset?
Yes, I have created a new variable called 'binary.quality' to determine if wine 
is good or bad. Wines that have a quality rating of 6 or higher are deemed good.
Any wine with a rating less than 6 is considered bad. Creating these two groups 
will help to uncover patterns from a bi-variate and multivariate analysis.

```{r echo=FALSE}
#Label wines as good or bad based on quality rating
cond <- pd$quality >= 6
pd$binary.quality <- ifelse(cond, 'Good', 'Bad')

```


```{r echo=FALSE,message=FALSE,warning=FALSE}
ggplot(data = pd, aes(x = binary.quality)) + geom_histogram(stat="count") + 
  xlab('Quality') + ylab('Count') + ggtitle("Quality of Red Wine's in Dataset")

ggplot(data = pd, aes(x = binary.quality, y = volatile.acidity)) +
  geom_boxplot()
```

### Of the features you investigated, were there any unusual distributions? 
In volatile acidity, it appears to be a bi modal distribution. There are 
multiple outliers in this distribution where the wines have volatile acidities 
higher than 1.015. No, I have not performed any data adjustment. 


# Bivariate Plots Section

```{r echo=FALSE,Bivariate_Plots_pair.plot}
ggpairs(pd, columns = 2:13, columnLabels=c('fixed.acidity','volatile.acidity',                                           'citric.acid','residual.sugar','chlorides','free.sulfur.dioxide',
    'total.sulfur.dioxide','density','pH','sulphates','alchohol','quality')  ,
        upper=list(continuous = wrap("cor", size=3), axisLabels='show'))




```

A pair-plot of the variables in the dataset reveals that alcohol, volatile 
acidity, and citric acid seem to have a strong relationship with the quality of 
a wine. I will explore these relationships further. 

```{r echo=FALSE,Bivariate_Plots_fixed.acidity}
ggplot(data = pd, aes(x = binary.quality, y = volatile.acidity, 
    fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_fill_brewer(palette = "Set1") + 
  ggtitle('Volatile Acidity by Wine Quality') + xlab("Wine Quality") + 
  ylab("Volatile Acidity") + labs(fill = 'Quality')  + 
  theme(plot.title = element_text(hjust = 0.5))

by(pd$volatile.acidity, pd$binary.quality, summary)

```

The boxplot above shows that the variance for volatile acidity in good and bad 
wines is similar. Bad wines have a higher mean and median. The mean and median 
volatile acidity for good wines was 0.4741 and 0.46 respectively. The mean and 
median volatile acidity for bad wines was 0.5895 and 0.59 respectively. 

```{r echo=FALSE,Bivariate_Plots_sulphates}
ggplot(data = pd, aes(x = binary.quality, y = sulphates, 
      fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  +
  scale_fill_brewer(palette = "Set1") + 
  ggtitle('Sulphate level by Wine Type') + 
  xlab("Wine Quality") + ylab("Sulphates") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(pd$sulphates, pd$binary.quality, summary)

```

Sulphate levels also vary between wines of different qualities. Good quality 
wines had higher mean and median sulphate levels. The mean and median values for
good wines were 0.6926 and 0.66 respectively. Wines of poor quality had mean and
median sulphate levels of 0.6185 and 0.58 respectively. 

```{r echo=FALSE,Bivariate_Plots_alcohol}
ggplot(data = pd, aes(x = binary.quality, y = alcohol, 
      fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_y_continuous(breaks = seq(8, 14, 1)) + 
  coord_cartesian(ylim = c(8, 14)) + 
    scale_fill_brewer(palette = "Set1") + 
  ggtitle('Alcohol Content by Wine Type') + xlab("Wine Quality") + 
  ylab("Alcohol Content") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(pd$alcohol, pd$binary.quality, summary)

```

Alcohol content varies depending on the quality of the wine. Wines of good 
quality have a higher variance as well as a higher mean and median. The mean and
median alcohol content for good wines was 10.86 and 10.8 respectively. Bad wines
had a mean alcohol content of 9.926 and a median of 9.7. 

```{r echo=FALSE,Bivariate_Plots_total.sulfur.dioxide}
ggplot(data = pd, aes(x = binary.quality, y = total.sulfur.dioxide, 
    fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_y_continuous(breaks = seq(0, 180, 40)) + 
  coord_cartesian(ylim = c(6, 180)) + scale_fill_brewer(palette="Set1") + 
  ggtitle('Total Sulfur Dioxide by Wine Type') + xlab("Wine Quality") +
  ylab("Total Sulfur Dioxide") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(pd$total.sulfur.dioxide, pd$binary.quality, summary)


```
Wines of poor quality have a larger variance in total sulfur dioxide levels as 
well as a higher mean and median. Wines of bad quality have a median of 45 and a
mean of 54.65. Wines of good quality have a mean of 39.35 and a median of 33. 

```{r echo=FALSE,Bivariate_Plots_citric.acid}
ggplot(data = pd, aes(x = binary.quality, y = citric.acid, 
    fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_y_continuous(breaks = seq(0, 1, .1)) + coord_cartesian(ylim = c(0, 1)) + 
    scale_fill_brewer(palette ="Set1") + 
  ggtitle('Citric Acid by Wine Quality') + xlab("Quality") + 
  ylab("Citric Acid") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))

by(pd$citric.acid, pd$binary.quality, summary)


```

A box plot of the citric acid content of wines by quality reveals that good 
wines have a higher median and variance than bad wines. Good wines also have a 
higher mean citric acid. The mean and median levels for good wines are 0.46 and 
0.31 respectively. Bad wines have a mean of 0.2378 and a median of 0.22. 

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
##investigation. How did the feature(s) of interest vary with other features in 
##
the dataset?

```{r echo=FALSE}
ggplot(data = pd, aes(x = volatile.acidity, y = alcohol)) +
  geom_point(alpha = 1/4, color = 'darkblue') + 
  ggtitle('Alcohol vs. Fixed Acidity') + xlab("Fixed Acidity") + ylab("Alcohol")

cor(pd$alcohol, pd$fixed.acidity)
```

A bi-variate analysis of the dataset reveals that good wines have tend to have a
higher alcohol content as well as lower levels of volatile acidity. These two 
variables do not seem to have a significant correlation with each other, but I 
will explore this relationship in more detail in my multivariate analysis.



### Did you observe any interesting relationships between the other features 
###(not the main feature(s) of interest)?

One interesting relationship that wasn't a main feature of interest, is that of 
fixed acidity and density. These two variables seem to have a significant 
correlation. This will be explored further. 

### What was the strongest relationship you found?

```{r echo=FALSE}
cor(pd$pH, pd$fixed.acidity)
```

Fixed acidity and pH have the strongest correlation among all the variables 
analyzed in this dataset. The correlation coefficient for these two variables is
-0.6829782. This would make sense because the pH scale is a measure of how basic
or acidic a substance is. One would expect a high correlation for these two 
variables. 

# Multivariate Plots Section

```{r echo=FALSE,Multivariate_Plots_fixed.acidity.v.density}
ggplot(data = pd, aes(x = fixed.acidity, y = density)) +
  geom_point(aes(color = binary.quality)) + 
  scale_x_continuous(breaks = seq(4, 16, 1)) + 
  scale_y_continuous(breaks = seq(0.99, 1.1, .002)) +
ggtitle('Fixed Acidity vs. Density') + xlab("Fixed Acidity") + ylab("Density") +
  theme(plot.title = element_text(hjust = 0.5))

cor(pd$fixed.acidity, pd$density)
```

In the previous section, I noted that density and fixed acidity seem to have a 
significant correlation. I examined this relationship further by creating a 
scatter plot of the two variables and labeling each point based on the quality 
of the underlying wine. It seems like at lower fixed acidity levels, good wines 
tend to have lower densities. 

```{r echo=FALSE,Multivariate_Plots}
ggplot(data = pd, aes(x = alcohol, y = volatile.acidity)) + 
  geom_point(aes(color = binary.quality))
```

It seems like the higher the alcohol content and lower the volatile acidity of a
wine, the more likely it is to be of good quality. 

```{r echo=FALSE}
ggplot(data = pd, aes(x = alcohol, y = sulphates)) + 
  geom_point(aes(color = binary.quality))
```

Higher sulphate levels and alcohol content also seem to be indicators of a good 
wine.

```{r echo=FALSE}
ggplot(data = pd, aes(x = alcohol, y = citric.acid)) + 
  geom_point(aes(color = binary.quality))
```

Similarly, wines with higher alcohol and citric acid levels tend to be of good 
quality.


# Multivariate Analysis
Wines with higher alcohol, volatile acidity, and sulphate levels tend to be of 
better quality. One could use these features to construct a linear model to 
predict the quality of a red wine. 


# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(data = pd, aes(x = binary.quality, y = volatile.acidity, 
                        fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_fill_brewer(palette = "Set1") + 
  ggtitle('Volatile Acidity by Wine Quality') + xlab("Wine Quality") + 
  ylab("Volatile Acidity (g / dm^3)") + labs(fill = 'Quality')  + 
  theme(plot.title = element_text(hjust = 0.5))

```

Most wines with a volatile acidity less than 0.6 tend to be of good quality. 
Wines with a volatile acidity higher than 0.6 were more likely to be of bad 
quality

### Plot Two
### Alcohol Content and Wine Quality

```{r echo=FALSE,Plot_Two}
ggplot(data = pd, aes(x = binary.quality, y = alcohol, 
      fill = binary.quality)) + 
  geom_boxplot(outlier.colour = "red", outlier.alpha = 1/4)  + 
  scale_y_continuous(breaks = seq(8, 14, 1)) + 
  coord_cartesian(ylim = c(8, 14)) + 
    scale_fill_brewer(palette = "Set1") + 
  ggtitle('Alcohol Content by Wine Type') + xlab("Wine Quality ") + 
  ylab("Alcohol Content (% by volume)") + labs(fill = 'Quality')  +
  theme(plot.title = element_text(hjust = 0.5))
```

Most wines with an alcohol content greater than 10% are of good quality. There 
are significantly more wines with alcohol contents between 9 and 10% that are of 
bad quality. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(data = pd, aes(x = volatile.acidity, y = alcohol)) + 
  geom_point(alpha = 1/3, aes(color = binary.quality)) + 
  scale_x_continuous(breaks =seq(0, 1.5, .2))  + 
  scale_y_continuous(breaks  = seq(0, 16, 2)) +   
  ggtitle('Alcohol Content vs. Volatile Acidity by Wine Quality') +
  xlab("Volatile Acidity (g / dm^3)") +
  ylab("Alcohol (% by volume)") + labs(color="Quality") + theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5), panel.border = element_blank(), 
        panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), 
axis.line = element_line(colour = "green")) + 
  scale_colour_manual(values = c("#A60000", "#0047A6"))

```

Wines with higher alcohol contents and lower volatile acidity levels were more 
likely to be of good quality. Most wines with alcohol contents greater than 10% 
and volatile acidity levels lower than 0.6 (g / dm^3) are of good quality.


# Reflection

Exploring the wine dataset revealed a relationship between the volatile acidity,
alcohol content, and quality of a wine. Prior to conducting a bi-variate and 
multivariate analysis, the wines were divide into two groups(good/bad). Encoding
the wines with a binary variable helped to discern which features had the most 
influence over the quality of wines. The higher the alcohol content of a wine, 
the more likely it is of good quality. Good quality wines also have lower 
volatile acidity levels.

Initially when I was developing for this project, I noticed that some wines 
didn't have citric acid at all. My first thought was maybe this was bad data or 
incomplete data. But then I researched further about wines. I saw that citric 
acid actually is added to some wines to increase the acidity.So it's evident 
that some wines would not have Citric Acid at all. So actually 
this was in parallel to my experimental findings.

For future analysis, I would love to have a dataset, where apart from the wine 
quality, a rank is given for that particular wine by 5 different wine tasters 
as we know when we include the human element, our opinion changes on so 
many different factors. So by including the human element in my analysis, 
I would be able to put in that perspective and see a lot of unseen factors 
which might result in a better or worse wine quality. Having these factors 
included inside the dataset would result in a different insight altogether 
in my analysis.